Signal Processing to Demodulate FM Radio

### 2020-12-16 doc/paper/paper.tex
almost entirely done with it

  \subsection{Frequency Discriminator}
    A frequency discriminator can be defined as a black box component, that generates an output that is directly proportional to the frequency of the input FM signal.
    Different strategies can be implemented in this black box component, the discriminator.
    Here, a differentiator with a subsequent envelope detector is used, which is shown in Fig.\ref{fig_freq_discriminator_bd}.

    \begin{figure}[!h]
      \centering
        \includegraphics[width=7cm]{img/discrimination_method_bd1.jpg}
      \caption{Frequency Discriminator Block Diagram \cite{ref_fig_freq_discriminator}}
      \label{fig_freq_discriminator_bd}
    \end{figure}

    This type of demodulator converts the FM signal into an AM signal.
    However, this signal still contains an FM content.
    The transformation that happens here can be expressed in a formula, by simply differentiating equation \ref{equ_fm_func}.
    This operation, with simple algebraic transformations, results in
    \begin{equation}
      \frac{d s(t)}{dt} = A_c 2 \pi \Big[f_c + k_f m(t) \Big] sin \Big(\omega_c t + 2 \pi k_f \int m(t) dt -pi \Big)
    \end{equation}

@@

    It can clearly be seen, that the signal now contains an AM and an FM content.
    The FM part is described by the sine function.
    More importantly the AM part, which resembles the envelope of the signal, is described by the term within the square brackets.
    The resulting signal in time-domain is shown in the top left diagram in Fig.\ref{fig_time_domain_envelope_detect}.
    This signal needs to be fed into an envelope detector, as shown in Fig.\ref{fig_freq_discriminator_bd}.
    It extracts the envelope by removing the high frequency portion.
    The method implemented here is one of the most simple ones and is usually referred to as 'Asynchronous Half-Wave Envelope Detector'\cite{ref_envelope_detector}.
    It consists of a thresholding unit to remove negative values and a conventional lowpass filter.
    In analog implementations, the thresholding unit is a diode.
    The lowpass' cutoff frequency needs to be adapted to the envelope signals' maximum frequency, since it needs to be able to follow the signal, but also sufficiently smooth out the rectified signal.
    Different methods may be implemented for the envelope detector.
    For example, an 'Asynchronous Full-Wave Envelope Detector' can be used.
    Therefore, the mentioned method only needs to swap the thresholding unit with an unit that calculates an absolut value.
    In that architecture, a full-wave rectification is performed on the signal, which significantly improves the envelope detection accuracy \cite{ref_envelope_detector}.
    The envelope is followed more exactly, because of the higher power that is available in the signal, since the entire signal is taken into account, and not only the positive half, as in the previous method.

    \begin{figure}[!h]
      \centering
        \includegraphics[width=8.8cm]{img/envelope-detect-time-domain.jpg}
      \caption{Time-Domain Signals in Envelope Detection \cite{ref_time_domain_envelope_detect}}
      \label{fig_time_domain_envelope_detect}
    \end{figure}

    An important requirement for the differentiator is, that the input signal is of a constant amplitude \cite{ref_schnyder_haller}.
    The transmitted signal is subject to random additive noise on the transmission channel.
    The differentiator would deliver wrong results because of this noise, when substracting two consecutive samples, as described in Equation \ref{equ_differentiator_two_samples}.
    To achieve this, the complex signal $s(t)$ can be normalized to a value of one.
    \begin{equation}
      s_{norm} = \frac{s}{|s|} = \frac{A(n)\ e^{j\phi_{FM}(n)}}{|{A(n)\ e^{j\phi_{FM}(n)}|}} = e^{j\phi_{FM}(n)} = |1|
    \end{equation}

@@

  \subsection{Mono}
    The first part of the channel frequency spectrum is the summation signal of the left and right audio channel.
    A mono receiver can thus simply replay the so-called multiplex signal, which is generated by the FM demodulation described in chapter \ref{sec_fm_sig_demod}. The 19 kHz pilot tone will not be audible by most people, because it is outside of the range of a human ear. Besides that, any used speaker may also be unable to produce a frequency in this range. This explanation applies to several higher frequency components in the spectrum as well.
    In order to be able to decimate the signal to a lower samplerate, for example to store it in a file with 44.1 kHz samplerate, a lowpass needs to be inserted before the decimation and output.
    The lowpass' cutoff frequency needs to be configured after 15 kHz.
    The filter needs to reach a sufficient attenuation before the pilot tone at 19 kHz.

  \subsection{Stereo}
    Demodulating an FM channel to recover a stereo audio signal, requires a more sophisticated approach.
    In the channel frequency spectrum, there are two main parts of the audio signal, the left-right sum and -difference signals.
    To combine these to a stereo signal, the following equations can be applied.
    \begin{equation*}
      L = (L+R) + (L-R) = (2)L
    \end{equation*}
    \begin{equation*}
      R = (L+R) - (L-R) = (2)R
    \end{equation*}

  % page 30: https://www.hs-schmalkalden.de/fileadmin/portal/Dokumente/Fakult%C3%A4t_ET/Personal/Roppel/Buch/Analoge_Modulationsverfahren.pdf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Matlab Simulation}
  % TODO: insert entire block diagram of demodulation
  %       HF => abs() => discriminator => AM demod => audio

  \subsection{Data Acquisition}
    % RTL2832u dongle
    % Linux driver
    % IQ interleaved data stream

  \subsection{Demodulation Steps}
    \subsubsection{Recorded IF spectrum}
    \subsubsection{Modulate to FM channel center frequency}
    \subsubsection{Decimation}
    \subsubsection{Demodulation}
    \subsubsection{Verification}
      % is done subjectively only, by listening to the audio stream via a PCs' soundcard.

\section{Conclusion}
  The conclusion goes here.

@@

  \bibitem{ref_time_domain_envelope_detect}
    C.Roppel, \emph{Analoge Modulationsverfahren und Rundfunktechnik} Begleitmaterial zum Buch 'Grundlagen der digitalen Kommunikationstechnik, Übertragungstechnik - Signalverarbeitung - Netze', \url{https://www.hs-schmalkalden.de/fileadmin/portal/Dokumente/Fakult%C3%A4t_ET/Personal/Roppel/Buch/Analoge_Modulationsverfahren.pdf}, opened on Dec. 16th, 2020

>>>>

  \subsection{Frequency Discriminator}
    A frequency discriminator can be defined as a black box component, that generates an output that is directly proportional to the frequency of the input FM signal.
    Different strategies can be implemented in this black box component, the discriminator.
    Here, a differentiator with a subsequent envelope detector is used, which is shown in Fig.\ref{fig_bd_freq_discriminator}.

    \begin{figure}[!h]
      \centering
        \includegraphics[width=7cm]{img/discrimination_method_bd1.jpg}
      \caption{Frequency discriminator block diagram \cite{ref_fig_freq_discriminator}}
      \label{fig_bd_freq_discriminator}
    \end{figure}

    This type of demodulator converts the FM signal into an AM signal.
    However, this signal still contains an FM content.
    The transformation that happens here can be expressed in a formula, by simply differentiating Eq.\ref{equ_fm_func}.
    This operation, with simple algebraic transformations, results in
    \begin{equation}
      \frac{d s(t)}{dt} = A_c 2 \pi \Big[f_c + k_f m(t) \Big] sin \Big(\omega_c t + 2 \pi k_f \int m(t) dt -pi \Big)
    \end{equation}

@@

    It can clearly be seen, that the signal now contains an AM and an FM content.
    The FM part is described by the sine function.
    More importantly the AM part, which resembles the envelope of the signal, is described by the term within the square brackets.
    The resulting signal in time-domain is shown in the top left diagram in Fig.\ref{fig_time_domain_envelope_detect}.
    This signal needs to be fed into an envelope detector, as shown in Fig.\ref{fig_bd_freq_discriminator}.
    It extracts the envelope by removing the high frequency portion.
    The method implemented here is one of the most simple ones and is usually referred to as 'Asynchronous Half-Wave Envelope Detector'\cite{ref_envelope_detector}.
    It consists of a thresholding unit to remove negative values and a conventional lowpass filter.
    In analog implementations, the thresholding unit is a diode.
    The lowpass' cutoff frequency needs to be adapted to the envelope signals' maximum frequency, since it needs to be able to follow the signal, but also sufficiently smooth out the rectified signal.
    Different methods may be implemented for the envelope detector.
    For example, an 'Asynchronous Full-Wave Envelope Detector' can be used.
    Therefore, the mentioned method only needs to swap the thresholding unit with an unit that calculates an absolut value.
    In that architecture, a full-wave rectification is performed on the signal, which significantly improves the envelope detection accuracy \cite{ref_envelope_detector}.
    The envelope is followed more exactly, because of the higher power that is available in the signal, since the entire signal is taken into account, and not only the positive half, as in the previous method.

    \begin{figure}[!h]
      \centering
        \includegraphics[width=8.8cm]{img/envelope-detect-time-domain.jpg}
      \caption{Time-domain signals in envelope detection \cite{ref_roppel}}
      \label{fig_time_domain_envelope_detect}
    \end{figure}

    An important requirement for the differentiator is, that the input signal is of a constant amplitude \cite{ref_schnyder_haller}.
    The transmitted signal is subject to random additive noise on the transmission channel.
    The differentiator would deliver wrong results because of this noise, when substracting two consecutive samples, as described in Eq.\ref{equ_differentiator_two_samples}.
    To achieve this, the complex signal $s(t)$ can be normalized to a value of one.
    \begin{equation}
      s_{norm} = \frac{s}{|s|} = \frac{A(n)\ e^{j\phi_{FM}(n)}}{|{A(n)\ e^{j\phi_{FM}(n)}|}} = e^{j\phi_{FM}(n)} = |1|
    \end{equation}

@@

  \subsection{Mono}
  \label{demod_mono}
    The first part of the channel frequency spectrum is the summation signal of the left and right audio channel.
    A mono receiver can thus simply replay the so-called multiplex signal, which is generated by the FM demodulation described in chapter \ref{sec_fm_sig_demod}. The 19 kHz pilot tone will not be audible by most people, because it is outside of the range of a human ear. Besides that, any used speaker may also be unable to produce a frequency in this range. This explanation applies to several higher frequency components in the spectrum as well.
    In order to be able to decimate the signal to a lower samplerate, for example to store it in a file with 44.1 kHz samplerate, a lowpass needs to be inserted before the decimation and output.
    The lowpass' cutoff frequency needs to be configured after 15 kHz.
    The filter needs to reach a sufficient attenuation before the pilot tone at 19 kHz.

  \subsection{Stereo}
    Demodulating an FM channel to recover a stereo audio signal requires a more sophisticated approach.
    In the channel frequency spectrum, there are two main parts of the audio signal - the sum and difference of the left and right channel audio signals.
    To combine these to a stereo signal, the following equations can be applied.
    \begin{equation*}
      L = (L+R) + (L-R) = (2)L
    \end{equation*}
    \begin{equation}
      R = (L+R) - (L-R) = (2)R
      \label{equ_stereo_from_sum_diff}
    \end{equation}

    The block diagram in Fig.\ref{fig_bd_stereo_demod} illustrates the signal processing chain for stereo audio. In the block diagram, $x(t)$ represents the multiplex signal, which comes from the FM demodulator. Signals $x_l(t)$ and $x_r(t)$ describe the left and right audio signal, respectively.

    \begin{figure}[!h]
      \centering
        \includegraphics[width=8.8cm]{img/fm-demod-stereo-audio.png}
      \caption{Block diagram of FM stereo audio demodulation \cite{ref_roppel}}
      \label{fig_bd_stereo_demod}
    \end{figure}

    The central branch selects the pilot tone at 19 kHz with a bandpass filter.
    This bandpass needs to have a frequency response that is sharp enough to have a sufficient attenuation at $\pm$4 kHz, since this is where the summation and difference signals frequencies are allocated.
    Afterwards, the pilot tone frequency is doubled via a PLL to achieve a 38 kHz to later demodulate the difference signal phase-coherently.
    The PLL lock indicator can be re-used as an indicator, whether a pilot tone is existent.
    In the upper branch, a bandpass filter selects the difference signal with ranges from 23 to 53 kHz.
    It is then modulated to baseband by the previously generated 38 kHz subcarrier.
    Subsequently, a lowpass limits the signal to the audio signal bandwidth of 15 kHz.
    The lower branch lowpass-filters the summation signal.
    The rightmost part in the diagram performs the combination of summation and difference signals, according to the Eq.\ref{equ_stereo_from_sum_diff}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Matlab Simulation}
  Matlab is used to simulate an FM demodulator.
  The objective is to demodulate a previously recorded data stream, so that one can listen to a mono audio signal.
  Several steps that are required therefore are described briefly in the following sections.

  % TODO: insert entire block diagram of demodulation
  %       HF => abs() => discriminator => AM demod => audio

  \subsection{Data Acquisition}
    A data stream is recorded, using a cheap RTL-SDR dongle.
    The device can be plugged into USB on a host computer.
    It features an RTL2832U chipset.
    In combination with a host software, this basically assembles a software defined radio (SDR).
    The recorded data stream is stored as a file with interleaved IQ baseband data.

  \subsection{Demodulation Steps}
    The SDR is used to record a data stream with a sampling frequency of 1 MHz, at a center frequency of 98.0 MHz. A dataset of 10 seconds is recorded.
    The aim is to listen to the Austrian radio station OE3, which is broadcasted at a center frequency of 98.1 MHz in the region of Saalfelden.

    \subsubsection{Recorded spectrum}
      The file was recorded at a center frequency $f_c$ of 98 MHz, while the desired station is located at 98.1 MHz.
      This means, that the recorded spectrum is allocated at an intermediate frequency (IF) of 0.1 MHz.

    \subsubsection{Modulate to FM channel center frequency}
      Because the signal needs to be in baseband, to be able to demodulate it, the spectrum needs to be shifted by 0.1 MHz.
      This is achieved by modulating with a complex signal with exactly that frequency.

    \subsubsection{Decimation}
      The signal still has a sample rate of 1 MHz, which is an overkill for the application, since a channels' bandwith only ranges up to 100 kHz.
      Keeping in mind the Nyquist theorem, the sample rate therefore needs to be at least 200 kHz.
      Thus, the signal is decimated by an arbitrary factor of 4 % TODO: more? see Nyquist.. work this out in script
      , which still leaves a rate of x M/kHz.

    \subsubsection{Demodulation}
      Demodulation is done by normalizing the signal amplitude to one.
      A FIR filter is used as a differentiator.
      Envelope detection is omitted for now, for simplicity and because of the mentioned effects in section \ref{demod_mono}.
    \subsubsection{Verification}
      Verification of correct functionality is only done subjectively, by listening to the demodulated audio signal via the PCs' soundcard.
      The OE3 stations' signal can be heard clearly.

\section{Conclusion}
  FM is a well-known technology, that is established in anyones daily life.
  Looking at it from a signal processing perspective is an interesting task, since it requires many principles of digital signal processing, but is still comparably simple to demodulate.

@@

  \bibitem{ref_roppel}
    C.Roppel, \emph{Analoge Modulationsverfahren und Rundfunktechnik} Begleitmaterial zum Buch 'Grundlagen der digitalen Kommunikationstechnik, Übertragungstechnik - Signalverarbeitung - Netze', \url{https://www.hs-schmalkalden.de/fileadmin/portal/Dokumente/Fakult%C3%A4t_ET/Personal/Roppel/Buch/Analoge_Modulationsverfahren.pdf}, opened on Dec. 16th, 2020
